### 2016-06-21 ansi-isolation.tex
edits

Note that, unlike recent proposals (\emph{e.g.},
~\cite{gotsmanconcur15}, our specifications for {\sc si} and {\sc ser} do not
necessarily impose a total order among transactions (conflicting or
otherwise). In reality, a total order under {\sc ser} (resp. {\sc si}) is guaranteed
only if the store executes all transactions under {\sc ser} (resp. {\sc SI})
isolation. Our specifications admit this possibility, and derive a total
order under the assumption of homogenity. However, databases almost
always allow isolation levels to be configured on a per-transaction
basis, allowing transactions at various isolation levels to coexist.
Specifications that assume homogenity are incorrect under this
setting.

>>>>

In contrast to recent proposals (\emph{e.g.},
~\cite{gotsmanconcur15}), our specifications for {\sc si} and {\sc
  ser} do not necessarily impose a total order among transactions
(conflicting or otherwise). In reality, a total order under {\sc ser}
(resp. {\sc si}) is guaranteed only if the store executes all
transactions under {\sc ser} (resp. {\sc SI}) isolation. Our
specifications admit this possibility, and derive a total order under
the assumption of homogenity. However, databases almost always allow
isolation levels to be configured on a per-transaction basis, allowing
transactions at various isolation levels to coexist.  Specifications
that assume homogenity are incorrect under this setting.

### 2016-06-21 store-consistency.tex
edits

Under a trivial trace invariant (i.e., $\I(\E) = true$), the abstract
machine of Fig.~\ref{fig:txnimp} assumes the semantics of an EC store;
it allows operations of a transaction witness arbitrary subsets of the
global state. A non-trivial $\I$ composed of isolation specifications
from Fig.~\ref{fig:ansi-isolation} induces the machine to provide
non-trivial isolation guarantees for transactions. However, weak
isolation levels often only constrain the visibility sets of
operations by dictating what \emph{not} to see; not what to see.  For
instance, \iso{Repeatable Read} isolation prohibits operations of a
transaction from witnessing different states. It however does not
prohibit all operations of a transaction from witnessing an aribitrary
subset of the global state. Consequently, the machine can remain an EC
store even while providing non-trivial isolation. How then to model
the semantics of an SC store, such as a relational database, with
variable isolation?

The answer lies in enforcing store-specific consistency constraints,
along with transaction-specific isolation constraints, via the trace
invariant $\I$. In particular, we split $\I$ into two components: (1).
$\I_s$, the store-specific invariant, and (2). $\I_c$, the
program-specific (or, client-specific) invariant, to capture
consistency and isolation constraints, respectively. $\I$ is now a
conjunction:
\begin{smathpar}
  \I \;=\; \lambda\E.~\I_s(\E) \wedge \I_c(\E)
\end{smathpar}
The trace invariant we wrote in \S\ref{sec:ansi-isolation} for the
withdraw program of Fig.~\ref{fig:motiv-eg-1} now becomes its $\I_c$
and remains an invariant regardless of the store. $\I_s$ however
changes from store to store. We now consider various stores and
describe their corresponding $\I_s$.

\paragraph{An EC store} An EC store provides no additional consistency
guarantees besides those provided by the machine. Hence, its
$\I_s(\E)$ is $true$ for any $\E$.

\paragraph{An SC store} An SC store guarantees total order of all
operations w.r.t $\visZ$ that is consistent with their chronological
order. A straightforward $\I_s$ for this store is the SC guarantee
defined below:
\begin{smathpar}
\begin{array}{l}
  \C{SC}(\E) \;=\; \forall\eta_1,\eta_2.~\{\eta_1,\eta_2\}
  \subseteq \E.\A \conj \id(\eta_1) < \id(\eta_2) \\
  \hspace*{2in}\Rightarrow \underE{\eta_1 \visar \eta_2}
\end{array}
\end{smathpar}
Unfortunately, $\I_s=\C{SC}$ conflicts with all isolation
specifications of Fig.~\ref{fig:ansi-isolation}. For instance,
consider a case where $\I_c(\E) \;=\; \forall
T_i.~\underE{\C{RC}(T_i)}$. That is, all transactions execute under
\iso{Read Committed} isolation. Imagine a sample execution where
$\eta_1$'s transaction is not yet committed when $\eta_2$ is
generated. Letting $\eta_1$ be visible to $\eta_2$ violates $\I_c$,
whereas not letting it be visible violates $\I_{s}$. The only way to
satisfy both the invariants is to rule out all the executions that
interleave the operations of one transaction with the other, thereby
enforcing serializability and the ACID model of
transactions\footnote{Thus, serializability is a natural
generalization of SC to transactions.}. In general, when $\I_s$ is in
conflict (but not inconsistent) with $\I_c$, the only way to enforce
both is to restrict concurrency. Clearly, this is unacceptable since
it defeats the very purpose of weak isolation. 
% How then do we enforce weak isolation on a strongly consistent
% machine?

Relational databases demonstrate a way of out this impasse.
Implementations of weak isolation levels on relational databases
implicitly relax the SC requirement so as to maximize concurrency and
improve performance. Consequently, execution traces do not necessarily
satisfy \C{SC}. This approach hints at a general solution reconciling
$\I_s$ with $\I_c$, which is to weaken the former. However, there are
many ways $\I_s$ can be weakened; for instance, by setting it to
$true$, which makes the store EC. In reality though, weak isolation on
databases does not cause so drastic a reduction in consistency. It is
therefore possible that there exists a a principled approach to weaken
$\I_s$ that adequately captures the underlying reality. Such a
principle can be gleaned by observing that data stores often make
\emph{recency} commitments~\cite{bailishat}, by which they aim to make
the most recent data available to clients. SC is infact an extreme
form of a recency commitment, which guarantees that reads witness most
recent writes to a data item. In case this conflicts with weak
isolation guarantees, such as RR which requires all reads in a
transaction to witness the same state, the natural way to weaken
consistency while prioritizing recency is to let the reads witness
most recent state that \emph{does not} violate the isolation
constraints. In the context of the operational model of
Fig.~\ref{fig:txnimp}, this translates to making visible to an
operation the largest subset ($S$) of the global state ($\A$) that
does not lead to the violation of isolation constraints. Generalizing
this approach to any $\I_s$ and $\I_c$ gives the \emph{maximum
visiblity principle}, which requires the weakened consistency
guarantee ($\I_s'$) of a store to enforce all visibility relationships
imposed by the actual consistency guarantee ($\I_s$) on a given trace,
unless enforcing such a relationship violates $\I_c$.  Formally,
maximum visibility weakening of $\I_s$ is an $\I_s'$, such that:
\begin{itemize}
  \item $\I_s'$ is weaker than $\I_s$: 
      $\forall\E.~ \I_s(\E) \Rightarrow \I_s'(\E)$, and
  \item In every trace $\E$ that satisfies $\I_c$, and for every pair
  of effects $\eta_1$ and $\eta_2$ in $\E$, if $\I_s(\E)$ requires
  $\eta_1$ to be visible to $\eta_2$, then so does $\I_s'(\E)$ unless
  extending $\E$ with $\visZ(\eta_1,\eta_2)$ violates
  $\I_c$\footnote{\GK{ToDo: consider other well-formedness conditions
  on trace, such as acyclicity of $\visZ$ and $\soZ$. Are they needed
  (considering that the machine never violates them)? Encode the
  specifications in Z3 and make sure they are consistent.}}:
  \begin{smathpar}
  \begin{array}{l}
  \forall\E,\eta_1,\eta_2.~ \I_c(\E) \Rightarrow (\I_s(\E)
    \Rightarrow \underE{\eta_1 \visar \eta_2}) \Rightarrow \\
    \hspace*{0.5in}(\I_s'(\E) \Rightarrow \underE{\eta_1 \visar
    \eta_2} \disj \neg\I_c(\E\,\cup\,(\emptyset,\{(\eta_1,\eta_2)\})))
  \end{array}
  \end{smathpar}
\end{itemize}

Applying maximum visibility principle, we weaken \C{SC} to obtain the
following store trace invariant ($\I_s$) for an SC store whose
isolation constraints are captured by $\I_c$:
\begin{smathpar}
\begin{array}{lcl}
\I_s(\E) & = & \forall \eta_1,\eta_2.\, \{\eta_1,\eta_2\},
    \subseteq \E.\A \conj \id(\eta_1) <
    \id(\eta_2) \\
    & & \hspace*{0.5in} \Rightarrow 
      \underE{\eta_1 \visar \eta_2} \disj \neg\I_c(\E
    \cup (\emptyset,\{(\eta_1,\eta_2)\}))\\
\end{array}
\end{smathpar}
As usual $\I = \I_s \wedge \I_c$. The resultant operational model
hides an existing effect ($\eta\in\A$) from being visible to the
current operation if and only if showing it results in the violation
of $\I_c$.

\paragraph{A CC store} A causally consistent data store, such as the
one underlying the reasoning framework of~\cite{gotsmanpopl16}, allows
operations to only witness a causally consistent snapshot of the
global state. A straightforward trace invariant ($\I_s$) for this
store is the \C{CC} guarantee that requires an effect to witness all
causally preceding effects regardless of the isolation requirements of
its transaction:
\begin{smathpar}
\begin{array}{lccl}
\C{CC}(\E) & \;=\; &  & \forall \eta_1,\eta_2.\, 
      \E \Vdash \eta_1 \soar \eta_2 \Rightarrow  \underE{\eta_1 \visar
      \eta_2}\\
    &   & \wedge & \forall\eta_1,\eta_2,\eta_3.\,\underE{\eta_1 \visar
      \eta_2} \conj \underE{\eta_2 \visoar \eta_3} \\
    &   & &\hspace*{0.5in} \Rightarrow \underE{\eta_1 \visar \eta_3}\\
\end{array}
\end{smathpar}
However, like $\I_s = \C{SC}$, $\I_s = \C{CC}$ may induce the store to
meet isolation requirements by restricting concurrency, thus failing
to abide by the spirit of weak isolation. In order to admit weak
isolation without restricting concurrency, we weaken $\I_s$ while
being guided by the maximum visibility principle. The weakened $\I_s$
is shown below:
\begin{smathpar}
\begin{array}{lccl}
\C{CC}(\E) & \;=\; &  & \forall \eta_1,\eta_2.\, 
      \E \Vdash \eta_1 \soar \eta_2 \Rightarrow  \underE{\eta_1 \visar
      \eta_2} \\
    & & & \hspace*{0.6in}\disj \neg\I_c(\E \cup 
                (\emptyset,\{(\eta_1,\eta_2)\}))\\
    &   & \wedge & \forall\eta_1,\eta_2,\eta_3.\,\underE{\eta_1 \visar
      \eta_2} \conj \underE{\eta_2 \visoar \eta_3} \\
    &   & &\hspace*{0.3in} \Rightarrow \underE{\eta_1 \visar \eta_3}
      \disj \neg\I_c(\E \cup (\emptyset,\{(\eta_1,\eta_2)\}))\\
\end{array}
\end{smathpar}
Instantiating the parameter $\I$  with $\I_s \wedge \I_c$ in
Fig.~\ref{fig:txnimp} results in an operational semantics that admits
the violation of causal consistency if and only if the violation is
inevitable to enforce
$\I_c$.

>>>>

Under a trivial trace invariant (i.e., $\I(\E) = true$), the abstract
machine of Fig.~\ref{fig:txnimp} assumes the semantics of an {\sc ec} store;
it allows operations of a transaction to witness arbitrary subsets of the
global state. A non-trivial $\I$ composed of isolation specifications
from Fig.~\ref{fig:ansi-isolation} induces the machine to provide
non-trivial isolation guarantees for transactions. However, weak
isolation levels often only constrain the visibility sets of
operations by dictating what \emph{not} to see; not what to see.  For
instance, \iso{Repeatable Read} isolation prohibits operations of a
transaction from witnessing different states. It, however, does not
prohibit all operations of a transaction from witnessing an aribitrary
subset of the global state. Consequently, the machine can remain an {\sc ec}
store even while providing non-trivial isolation. How then to model
the semantics of an {\sc sc} store, such as a relational database, with
variable (weak) isolation?

The answer lies in enforcing store-specific consistency constraints,
along with transaction-specific isolation constraints, via the trace
invariant $\I$. In particular, we split $\I$ into two components: (1).
$\I_s$, the store-specific invariant, and (2). $\I_c$, the
program-specific (or, client-specific) invariant, to capture
consistency and isolation constraints, respectively. $\I$ is now a
conjunction:
\begin{smathpar}
  \I \;=\; \lambda\E.~\I_s(\E) \wedge \I_c(\E)
\end{smathpar}
The trace invariant defined in \S\ref{sec:ansi-isolation} for the
withdraw program of Fig.~\ref{fig:motiv-eg-1} now becomes its $\I_c$
and remains an invariant regardless of the store. $\I_s$ however
changes from store to store. We now consider various stores and
describe their corresponding store invariants.

\paragraph{An EC store} An {\sc ec} store provides no additional consistency
guarantees besides those provided by the machine. Hence, its
$\I_s(\E)$ is $true$ for any $\E$.

\paragraph{An SC store} An {\sc sc} store guarantees total order of all
operations w.r.t $\visZ$ that is consistent with their chronological
order. A straightforward $\I_s$ for this store is the {\sc sc} guarantee
defined below:
\begin{smathpar}
\begin{array}{l}
  \C{SC}(\E) \;=\; \forall\eta_1,\eta_2.~\{\eta_1,\eta_2\}
  \subseteq \E.\A \conj \id(\eta_1) < \id(\eta_2) \\
  \hspace*{2in}\Rightarrow \underE{\eta_1 \visar \eta_2}
\end{array}
\end{smathpar}
Unfortunately, $\I_s=\C{SC}$ conflicts with all isolation
specifications of Fig.~\ref{fig:ansi-isolation}. For instance,
consider a case where $\I_c(\E) \;=\; \forall
T_i.~\underE{\C{RC}(T_i)}$. That is, all transactions execute under
\iso{Read Committed} isolation. Imagine a sample execution where
$\eta_1$'s transaction is not yet committed when $\eta_2$ is
generated. Letting $\eta_1$ be visible to $\eta_2$ violates $\I_c$,
whereas not letting it be visible violates $\I_{s}$. The only way to
satisfy both the invariants is to rule out all the executions that
interleave the operations of one transaction with the other, thereby
enforcing serializability and the ACID model of
transactions\footnote{Thus, serializability is a natural
generalization of {\sc sc} to transactions.}. In general, when $\I_s$ is in
conflict (but not inconsistent) with $\I_c$, the only way to enforce
both invariant sets is to restrict concurrency. Clearly, this is unacceptable since
it defeats the very purpose of weak isolation. 
% How then do we enforce weak isolation on a strongly consistent
% machine?

Relational databases demonstrate a way of out this impasse.
Implementations of weak isolation levels on relational databases
implicitly relax the {\sc sc} requirement so as to maximize
concurrency and improve performance. Consequently, execution traces do
not necessarily satisfy {\sc sc}. This approach hints at a general
solution reconciling $\I_s$ with $\I_c$, that entails weakening the
former. However, there are many ways $\I_s$ can be weakened; for
instance, by setting it to $true$, which makes the store EC. In
reality though, weak isolation on databases does not cause so drastic
a reduction in consistency. It is therefore possible that there exists
a principled approach to weaken $\I_s$ that adequately captures the
underlying reality. Such a principle can be gleaned by observing that
data stores often make \emph{recency} commitments~\cite{bailishat}, by
which they aim to make the most recent data available to clients; {\sc
  sc} is, in fact, an extreme form of a recency commitment, which
guarantees that reads witness the most recent writes to a data
item. In case this conflicts with weak isolation guarantees, such as
{\sc rr} which requires all reads in a transaction to witness the same
state, the natural way to weaken consistency while prioritizing
recency is to let reads witness the most recent state that \emph{does
  not} violate the isolation constraints. In the context of the
operational model of Fig.~\ref{fig:txnimp}, this translates to making
the largest subset ($S$) of the global state ($\A$) that does not lead
to the violation of isolation constraints visible to an
operation. Generalizing this approach to any $\I_s$ and $\I_c$ yields
a \emph{maximum visiblity principle}, which requires the weakened
consistency guarantee ($\I_s'$) of a store to enforce all visibility
relationships imposed by the actual consistency guarantee ($\I_s$) on
a given trace, unless enforcing such a relationship violates $\I_c$.
Formally, the maximum visibility weakening of $\I_s$ is a new store
consistency relation $\I_s'$, such that:
\begin{itemize}
  \item $\I_s'$ is weaker than $\I_s$: 
      $\forall\E.~ \I_s(\E) \Rightarrow \I_s'(\E)$, and
  \item In every trace $\E$ that satisfies $\I_c$, and for every pair
  of effects $\eta_1$ and $\eta_2$ in $\E$, if $\I_s(\E)$ requires
  $\eta_1$ to be visible to $\eta_2$, then so does $\I_s'(\E)$ unless
  extending $\E$ with $\visZ(\eta_1,\eta_2)$ violates
  $\I_c$\footnote{\GK{ToDo: consider other well-formedness conditions
  on trace, such as acyclicity of $\visZ$ and $\soZ$. Are they needed
  (considering that the machine never violates them)? Encode the
  specifications in Z3 and make sure they are consistent.}}:
  \begin{smathpar}
  \begin{array}{l}
  \forall\E,\eta_1,\eta_2.~ \I_c(\E) \Rightarrow (\I_s(\E)
    \Rightarrow \underE{\eta_1 \visar \eta_2}) \Rightarrow \\
    \hspace*{0.5in}(\I_s'(\E) \Rightarrow \underE{\eta_1 \visar
    \eta_2} \disj \neg\I_c(\E\,\cup\,(\emptyset,\{(\eta_1,\eta_2)\})))
  \end{array}
  \end{smathpar}
\end{itemize}

Applying this principle, we can weaken {\sc sc} to obtain the
following store trace invariant ($\I_s$) for an {\sc sc} store whose
isolation constraints are captured by $\I_c$:
\begin{smathpar}
\begin{array}{lcl}
\I_s(\E) & = & \forall \eta_1,\eta_2.\, \{\eta_1,\eta_2\},
    \subseteq \E.\A \conj \id(\eta_1) <
    \id(\eta_2) \\
    & & \hspace*{0.5in} \Rightarrow 
      \underE{\eta_1 \visar \eta_2} \disj \neg\I_c(\E
    \cup (\emptyset,\{(\eta_1,\eta_2)\}))\\
\end{array}
\end{smathpar}
As usual $\I = \I_s \wedge \I_c$. The resultant operational model
hides an existing effect ($\eta\in\A$) from being visible to the
current operation if and only if showing it results in the violation
of $\I_c$.

\paragraph{A CC store} A causally consistent data store, such as the
one underlying the reasoning framework of~\cite{gotsmanpopl16}, allows
operations to only witness a causally consistent snapshot of the
global state. A straightforward trace invariant ($\I_s$) for this
store is the causally consistent guarantee that requires an effect to witness all
causally preceding effects regardless of the isolation requirements of
its transaction:
\begin{smathpar}
\begin{array}{lccl}
\C{CC}(\E) & \;=\; &  & \forall \eta_1,\eta_2.\, 
      \E \Vdash \eta_1 \soar \eta_2 \Rightarrow  \underE{\eta_1 \visar
      \eta_2}\\
    &   & \wedge & \forall\eta_1,\eta_2,\eta_3.\,\underE{\eta_1 \visar
      \eta_2} \conj \underE{\eta_2 \visoar \eta_3} \\
    &   & &\hspace*{0.5in} \Rightarrow \underE{\eta_1 \visar \eta_3}\\
\end{array}
\end{smathpar}
However, like $\I_s = \C{SC}$, $\I_s = \C{CC}$ we may induce the store to
meet isolation requirements by restricting concurrency, thus failing
to abide by the spirit of weak isolation. In order to admit weak
isolation without restricting concurrency, we weaken $\I_s$ while
being guided by the maximum visibility principle. The weakened $\I_s$
is shown below:
\begin{smathpar}
\begin{array}{lccl}
\C{CC}(\E) & \;=\; &  & \forall \eta_1,\eta_2.\, 
      \E \Vdash \eta_1 \soar \eta_2 \Rightarrow  \underE{\eta_1 \visar
      \eta_2} \\
    & & & \hspace*{0.6in}\disj \neg\I_c(\E \cup 
                (\emptyset,\{(\eta_1,\eta_2)\}))\\
    &   & \wedge & \forall\eta_1,\eta_2,\eta_3.\,\underE{\eta_1 \visar
      \eta_2} \conj \underE{\eta_2 \visoar \eta_3} \\
    &   & &\hspace*{0.3in} \Rightarrow \underE{\eta_1 \visar \eta_3}
      \disj \neg\I_c(\E \cup (\emptyset,\{(\eta_1,\eta_2)\}))\\
\end{array}
\end{smathpar}
Instantiating the parameter $\I$  with $\I_s \wedge \I_c$ in
Fig.~\ref{fig:txnimp} results in an operational semantics that admits
the violation of causal consistency if and only if the violation is
inevitable to enforce
$\I_c$.
